# The Visualization Framework

```{r include = FALSE}
knitr::opts_chunk$set(
   echo = FALSE
   , warning = FALSE
   , message = FALSE
   , error = FALSE
   , fig.width = 7
   , fig.height=3
)

library(tidyverse)
library(randomNames)
```

Having considered our audience, let's take a moment to talk about how data visualization actually comes into being. Most actuaries have generated a time series or bar chart and may find the process fairly straightforward. It is, but this simplicity belies the mathematical structure and set of decisions that underly the mapping of data to image.

```{r}
num_policies <- 5e3

set.seed(1234)
tbl_policy <- tibble(
  policyholder = randomNames(num_policies, which.names = 'last')
  , effective_date = as.Date('2001-01-01') + runif(num_policies, 0, 365)
  , expiration_date = effective_date + 365
  , premium = rnorm(num_policies, 10e3, sd = 1e3)
  , territory = sample(c('Athens', 'Sparta'), num_policies, replace = TRUE)
) %>% 
  mutate(
    month = lubridate::month(effective_date)
    , month = month.abb[month] %>% fct_relevel(month.abb)
  )
```

## The Tangibility of Data

Because we interact it with it so often, it may seem as though "data" is tangible. This sense comes via the mode through which we interact with it: a visual representation of data. This visubility comes about from the photons of light emitted by your laptop, the bits of ink on a sheet of paper, or the electric charge stored in a computer's memory. But these things are not actually the real-world elements, they are simply a representation of them. A "policyholder" is not a record in a database and would likely object to being referred to as such! 

Consider Table \@ref(tab:policy-table). This shows the first ten (of five thousand) records of a list of policyholders. Note that this is only a subset of the potential information which we might display. There could be claims, an address, demographics, marketing touch points, or any number of other characteristics. The data is no less real for being a subset of what we could choose to display.

```{r policy-table}
tbl_policy %>% 
  mutate(premium = format(premium, digits = 0, scientific = FALSE, big.mark = ",")) %>% 
  select(-month) %>% 
  select(territory, everything()) %>% 
  filter(nchar(policyholder) < 15) %>% 
  head(10) %>% 
  knitr::kable(
    align = 'clccr'
    , caption = "The first ten policyholders"
  )
```

We are accustomed to thinking of the set of recorded properties of real-world entities - like what is shown in Table \@ref(tab:policy-table) as being the true, accurate "data". Free yourself of that notion. Once you have taken that step, it will be easier to accept the authenticity of visualization as tangible "data".

## The visual space

When we create a visualization, we have a fixed space in which to place our data. It will be most useful to think of it as a rectangle, though this is not strictly required. Indeed, spherical coordinates, or certain mapping projections will result in non-rectangular plotting spaces. However, most visualizations may be expressed in this way.

A rectangular space gives us two dimensions to work with. That is it, just two. We can introduce additional information via attributes like color and shape, but there are only two physical dimensions. Returning to our policy data, we can see one possible representation of the dimensions of the visual space in Figure \@ref(fig:visual-space-empty). There is nothing within the space yet, we must begin by defining its size and scale.

```{r visual-space-empty, fig.cap = "An empty visual space"}
tbl_policy %>% 
  ggplot(aes(effective_date, premium))
```

We have naturally assumed that time should run along the horizontal, or $x$ axis and that premium should along the vertical, or $y$ axis. There is no requirement that we do so; it is simply a choice that we have made. The fact that it is sensible should not lead us to think that we are adhering to any particular obvious rule. It works, but we would not be violating any natural law by transposing the axes, as shown in Figure \@ref(fig:visual-space-transposed).

```{r visual-space-transposed, fig.cap = "Transposing the axes"}
tbl_policy %>% 
  ggplot(aes(premium, effective_date))
```

Take a moment and compare the linear distance available for the effective date in Figures \@ref(fig:visual-space-empty) and \@ref(fig:visual-space-transposed). Resist the temptation to think of that distance in temporal terms, and think only in physical terms. That is, think of distance as inches, centimeters, or pixels. There is more space in Figure \@ref(fig:visual-space-empty) than there is in Figure \@ref(fig:visual-space-transposed) - our rectangle is longer than it is wide. However, there is a precisely equal amount of _temporal space_. We are able to do this by relying on the computer's ability to scale the range of our data to the range of the axis. 

## Categorical axes

Physical distance is generally reckoned as coming from the set of real numbers. However, our plotting space is not bound by this. A dimension may be categorical - something which maps to the set of natural numbers - instead. The policy effective month is an obvious candidate for a categorial axis.

```{r visual-space-month, fig.cap = "Categorical data along the x-axis"}
tbl_policy %>% 
  ggplot(aes(month, premium))
```

There is often a natural preference for the order categories, but this, too, is something which the analyst must choose. In Figure \@ref(fig:visual-space-month), we have arranged the months in order, but we could have chosen something different. In Figure \@ref(fig:random-month), the months have been randomly sorted.

```{r random-month, fig.cap = "Months arranged randomly"}
tbl_policy %>% 
  mutate(month = fct_shuffle(month)) %>% 
  ggplot(aes(month, premium))
```

We could just as well have chosen `territory` as one of the axes. Consider what information could be shown in Figure \@ref(fig:territory-axis).

```{r territory-axis, fig.cap = "Another option for categorical data on the x-axis"}
tbl_policy %>% 
  ggplot(aes(territory, premium))
```

## Mapping 

The figures we have seen so far lack an obvious element: nothing appears in the plotting space. The process of expressing data visually requires a mapping of data to a visual element. We have already mapped two elements: the policy effective date (or the month) and the premium. In this case, we need do nothing more than inform our visual rendering engine what geometric shape to place in the plotting space. In Figure \@ref(fig:visual-space-point), we see the data represented as points. Each point corresponds to a single policyholder.

```{r visual-space-point, fig.cap = "A geometric shape as tangible data"}
tbl_policy %>% 
  ggplot(aes(effective_date, premium)) + 
  geom_point()
```

Figure \@ref(fig:visual-space-point) contains _more_ information than Table \@ref(tab:policy-table), yet consumes less physical space. One might argue that Table \@ref(tab:policy-table) contains more detail and this is true, to a point. It is valid to say that Table \@ref(tab:policy-table) would allow one to identify the specific premium and policy effective date for the first ten policyholders. If that's what the audience is looking for, you have succeeded in way that Figure \@ref(fig:visual-space-point) has not. However, as we learned in Section \@ref(sec-audience), this is likely not to be what your audience is looking for. What Figure \@ref(fig:visual-space-point) does do is answer the following questions:

* Is the business growing? 
* Is the business contracting?
* Is there any seasonality to our revenue?
* What is the average size of premium?
* Are there any policies which are particularly large or small?

We can use the same amount of visual space to introduce more information. As mentioned above, the two axis dimensions are already being used, but we could map a data element to color or shape. We see an example of this in Figure \@ref(fig:territory-color).

```{r territory-color, fig.cap = "Data represented via color"}
tbl_policy %>% 
  ggplot(aes(effective_date, premium)) + 
  geom_point(aes(color = territory))
```

In this section, we have shown some of the mechanics of bringing a data visualization into being. We have also shown the result of one particular set of choices. In the next section, we will explore more options for visual representation and consider how to choose among them.